---
title: "Comprehensive Abundance Index Standardization Workflow with standard methods"
subtitle: "From Data Collection to Model Validation"
author: "Sosthene AKIA"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
    number-sections: true
execute:
  warning: false
  message: false
---

# Introduction

## Overview

This document presents a **complete, production-ready workflow** for **abundance index standardization** following international best practices (Maunder & Punt 2004; ICES WGMG 2021).

We demonstrate the full process using **Pacific Cod** (*Gadus macrocephalus*) survey data from Queen Charlotte Sound, British Columbia (2003-2017).

## The 7-Step Standardization Framework

```{r framework_table, echo=FALSE}
library(knitr)

framework <- data.frame(
  Step = 1:7,
  Phase = c(
    "Data Collection & Cleaning",
    "Theoretical Model Design",
    "Model Implementation",
    "Model Selection & Validation",
    "Prediction Grid Construction",
    "Index Calculation",
    "Final Diagnostics & Reporting"
  ),
  Key_Activities = c(
    "Data import, QC, variable engineering, EDA",
    "Literature review, fishery knowledge, covariate selection",
    "Fit 12 candidate models (4 GLM, 4 GAM, 4 GLMM)",
    "AIC comparison, residual diagnostics, link function validation",
    "Define standardization strategy, build pred_data",
    "Predict with best models, calculate abundance index",
    "Spatial autocorrelation tests, compare models, finalize output"
  ),
  Tools = c(
    "tidyverse, ggplot2, summary tables",
    "Exploratory plots, correlation matrices, domain expertise",
    "glmmTMB, mgcv::gam, broom",
    "DHARMa, AIC, residual plots, gam.check()",
    "expand.grid(), reference values",
    "predict(), confidence intervals",
    "spdep::moran.test, sdmTMB, comparative plots"
  )
)

kable(framework, caption = "The 7-Step Framework for Index Standardization")
```

---

# STEP 1: Data Collection & Cleaning

## 1.1 Load Required Packages

```{r setup}
# Core packages
library(tidyverse)     # Data manipulation & visualization
library(sdmTMB)        # Pacific cod data + spatial models
library(mgcv)          # GAM models
library(glmmTMB)       # GLMM with Tweedie
library(DHARMa)        # Residual diagnostics
library(broom)         # Tidy model outputs
library(broom.mixed)   # Tidy mixed models
library(performance)   # Model comparison
library(patchwork)     # Combine plots
library(knitr)         # Tables

# Spatial diagnostics
library(spdep)         # Moran\'s I test
library(gstat)         # Variograms
library(sp)            # Spatial objects

# Set theme
theme_set(theme_minimal(base_size = 11))
```

## 1.2 Load Pacific Cod Data

```{r load_data}
# Load dataset
data("pcod", package = "sdmTMB")

# Initial inspection
cat("Dataset dimensions:", nrow(pcod), "rows ×", ncol(pcod), "columns\n\n")
glimpse(pcod)
```

**Dataset Description**:

- **Source**: Trawl survey in Queen Charlotte Sound, BC
- **Period**: 2003-2017 (15 years)
- **Sample size**: 2,143 trawl sets
- **Response**: `density` (kg/km²)
- **Covariates**: `year`, `depth`, `X`, `Y` (UTM coordinates)

## 1.3 Data Cleaning & Variable Engineering

```{r clean_data}
# Create analysis-ready dataset
dat_raw <- pcod |>
  select(year, X, Y, depth, density, present) |>
  filter(!is.na(density))

# Engineer variables
dat <- dat_raw |>
  mutate(
    # Response transformations
    log_density = log(density + 0.01),

    # Temporal variables
    fyear = as.factor(year),
    year_scaled = scale(year)[,1],

    # Depth variables (multiple transformations for model flexibility)
    depth_c = scale(depth)[,1],          # Centered & scaled
    depth_c2 = depth_c^2,                # Quadratic
    log_depth = log(depth + 1),          # Log-transformed
    sqrt_depth = sqrt(depth),            # Square root

    # Spatial variables
    X_c = scale(X)[,1],
    Y_c = scale(Y)[,1],

    # Regional indicator (North/South split)
    region = factor(ifelse(Y > median(Y), "North", "South")),

    # Interaction term
    region_year = interaction(region, fyear)
  )

# Summary statistics
cat("\n=== Data Summary ===\n")
cat("Total observations:", nrow(dat), "\n")
cat("Years:", min(dat$year), "-", max(dat$year), "\n")
cat("Depth range:", round(min(dat$depth)), "-", round(max(dat$depth)), "m\n")
cat("Zero catches:", sum(dat$density == 0), 
    "(", round(100*mean(dat$density == 0), 1), "%)\n")
cat("Median density:", round(median(dat$density[dat$density > 0]), 2), "kg/km²\n")
```

## 1.4 Data Quality Checks

```{r quality_checks, fig.height=8}
# 1. Check for outliers
p1 <- ggplot(dat, aes(x = density)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "A) Density Distribution", x = "Density (kg/km²)", y = "Count")

p2 <- ggplot(dat, aes(y = depth)) +
  geom_boxplot(fill = "coral", alpha = 0.7) +
  labs(title = "B) Depth Range", y = "Depth (m)")

# 2. Check temporal coverage
p3 <- ggplot(dat, aes(x = year)) +
  geom_bar(fill = "darkgreen", alpha = 0.7) +
  labs(title = "C) Sampling Effort by Year", x = "Year", y = "Number of Tows")

# 3. Check spatial coverage
p4 <- ggplot(dat, aes(x = X, y = Y)) +
  geom_point(alpha = 0.3, size = 0.5) +
  coord_equal() +
  labs(title = "D) Spatial Coverage", x = "X (km)", y = "Y (km)")

(p1 + p2) / (p3 + p4)
```

**Quality Assessment**:
- ✅ No extreme outliers detected
- ✅ Consistent sampling effort across years
- ✅ Good spatial coverage across survey area
- ⚠️ High proportion of zeros (Tweedie family appropriate)

---

# STEP 2: Theoretical Model Design

## 2.1 Exploratory Data Analysis (EDA)

### 2.1.1 Nominal Abundance Trend

```{r nominal_trend, fig.height=5}
# Calculate nominal index
nominal_index <- dat |>
  group_by(year) |>
  summarise(
    n = n(),
    mean_density = mean(density),
    se = sd(density) / sqrt(n()),
    median_density = median(density),
    .groups = "drop"
  )

ggplot(nominal_index, aes(x = year, y = mean_density)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_ribbon(aes(ymin = mean_density - 1.96*se,
                  ymax = mean_density + 1.96*se),
              alpha = 0.2, fill = "blue") +
  labs(title = "Nominal Abundance Index (Unstandardized)",
       subtitle = "Simple mean density per year",
       x = "Year", y = "Mean Density (kg/km²)") +
  scale_x_continuous(breaks = seq(2003, 2017, 2))
```

### 2.1.2 Covariate Relationships

```{r eda_covariates, fig.height=10}
# Density vs Depth (non-linear relationship expected)
p1 <- dat |>
  filter(density > 0) |>
  sample_n(min(1000, n())) |>
  ggplot(aes(x = depth, y = density)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 6), color = "red") +
  scale_y_log10() +
  labs(title = "A) Density vs Depth", x = "Depth (m)", y = "Density (log10)")

# Density by Region
p2 <- ggplot(dat, aes(x = region, y = log_density)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "B) Density by Region", x = "Region", y = "Log Density")

# Temporal trend by Region
p3 <- dat |>
  group_by(year, region) |>
  summarise(mean_dens = mean(density), .groups = "drop") |>
  ggplot(aes(x = year, y = mean_dens, color = region)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "C) Temporal Trends by Region", 
       x = "Year", y = "Mean Density (kg/km²)")

# Spatial distribution
p4 <- dat |>
  filter(year %in% c(2005, 2010, 2015)) |>
  ggplot(aes(x = X, y = Y, color = log(density + 0.1))) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap(~year) +
  scale_color_viridis_c(option = "plasma") +
  coord_equal() +
  labs(title = "D) Spatial Distribution (3 years)", 
       x = "X (km)", y = "Y (km)", color = "log(density)")

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = "Exploratory Data Analysis: Key Relationships")
```

**EDA Insights**:

1. **Depth effect**: Clear non-linear relationship → GAM with `s(depth)` warranted
2. **Regional differences**: North/South show different density levels → Include region
3. **Temporal variation**: Declining trend from 2003-2017
4. **Spatial structure**: Visible spatial clustering → Test spatial models

## 2.2 Model Design Decisions

Based on EDA and fishery knowledge:

```{r model_design_table, echo=FALSE}
design_decisions <- data.frame(
  Component = c(
    "Response distribution",
    "Temporal effect",
    "Depth effect",
    "Regional effect",
    "Spatial effect"
  ),
  Decision = c(
    "Tweedie (handles zeros + continuous)",
    "Year as factor (no trend assumption)",
    "Linear + Quadratic OR Spline",
    "Fixed effect OR Random effect",
    "No spatial OR 2D smooth OR Random field"
  ),
  Reasoning = c(
    "43% zeros, right-skewed positive values",
    "Flexible, no parametric trend assumption",
    "Test both parametric and non-parametric",
    "Compare parsimony vs flexibility",
    "Progressive complexity testing"
  )
)

kable(design_decisions, 
      caption = "Model Design Decisions Based on EDA & Domain Knowledge")
```

## 2.3 Candidate Model Set (12 Models)

We will fit **12 models** in 3 families:

### GLM Family (4 models)

```{r glm_formulas, echo=FALSE}
glm_models <- data.frame(
  Model = paste0("GLM", 1:4),
  Formula = c(
    "density ~ fyear + depth_c",
    "density ~ fyear + depth_c + depth_c2",
    "density ~ fyear + depth_c + depth_c2 + region",
    "density ~ fyear + depth_c + depth_c2 + region + year_scaled:region"
  ),
  Complexity = c("Minimal", "Baseline", "Regional", "Interaction")
)
kable(glm_models, caption = "GLM Model Set (glmmTMB, Tweedie)")
```

### GAM Family (4 models)

```{r gam_formulas, echo=FALSE}
gam_models <- data.frame(
  Model = paste0("GAM", 1:4),
  Formula = c(
    "density ~ fyear + s(depth, k=6)",
    "density ~ fyear + s(depth, k=6) + region",
    "density ~ fyear + s(depth, k=6) + s(X, Y, k=20)",
    "density ~ fyear + s(depth, k=6) + region + s(X, Y, k=20)"
  ),
  Complexity = c("Minimal", "Regional", "Spatial", "Full")
)
kable(gam_models, caption = "GAM Model Set (mgcv, Tweedie)")
```

### GLMM Family (4 models)

```{r glmm_formulas, echo=FALSE}
glmm_models <- data.frame(
  Model = paste0("GLMM", 1:4),
  Formula = c(
    "density ~ fyear + depth_c + depth_c2 + (1|region)",
    "density ~ fyear + depth_c + depth_c2 + (1|fyear:region)",
    "density ~ fyear + depth_c + depth_c2 + (1|region) +region:fyear",
    "density ~ fyear + depth_c + depth_c2 + (1|region) + (1|X:Y)"
))
kable(glmm_models, caption = "GLMM Model Set (glmmTMB, Tweedie)")
```

---

# STEP 3: Model Implementation

## 3.1 Fit GLM Models

```{r fit_glm_models}
cat("Fitting GLM models...\n")

# GLM1: Minimal
glm1 <- glmmTMB(
  density ~ fyear + depth_c,
  data = dat,
  family = tweedie(link = "log")
)

# GLM2: Baseline (quadratic depth)
glm2 <- glmmTMB(
  density ~ fyear + depth_c + depth_c2,
  data = dat,
  family = tweedie(link = "log")
)

# GLM3: + Region
glm3 <- glmmTMB(
  density ~ fyear + depth_c + depth_c2 + region,
  data = dat,
  family = tweedie(link = "log")
)

# GLM4: + Region interaction
glm4 <- glmmTMB(
  density ~ fyear + depth_c + depth_c2 + region + year_scaled:region,
  data = dat,
  family = tweedie(link = "log")
)

cat("✓ GLM models fitted\n")
```

## 3.2 Fit GAM Models

```{r fit_gam_models}
cat("Fitting GAM models...\n")

# GAM1: Minimal (spline depth)
gam1 <- gam(
  density ~ fyear + s(depth, k = 6),
  data = dat,
  family = tw(link = "log"),
  method = "REML"
)

# GAM2: + Region
gam2 <- gam(
  density ~ fyear + s(depth, k = 6) + region,
  data = dat,
  family = tw(link = "log"),
  method = "REML"
)

# GAM3: + Spatial smooth
gam3 <- gam(
  density ~ fyear + s(depth, k = 6) + s(X, Y, k = 20),
  data = dat,
  family = tw(link = "log"),
  method = "REML"
)

# GAM4: Full (region + spatial)
gam4 <- gam(
  density ~ fyear + s(depth, k = 6) + region + s(X, Y, k = 20),
  data = dat,
  family = tw(link = "log"),
  method = "REML"
)

cat("✓ GAM models fitted\n")
```

## 3.3 Fit GLMM Models

```{r fit_glmm_models, message=FALSE}
cat("Fitting GLMM models...\n")

# GLMM1: Random intercept (region)
glmm1 <- glmmTMB(
  density ~ fyear + depth_c + depth_c2 + (1|region),
  data = dat,
  family = tweedie(link = "log")
)

# GLMM2: Random slope (year|region)
glmm2 <- glmmTMB(
  density ~ fyear + depth_c + depth_c2 + (1|fyear:region),
  data = dat,
  family = tweedie(link = "log"),
  control = glmmTMBControl(optimizer = nlminb)
)

# GLMM3: 
glmm3 <- glmmTMB(
  density ~ fyear + depth_c + depth_c2 + (1|region) +region:fyear,
  data = dat,
  family = tweedie(link = "log")
)

# GLMM4: 
glmm4 <- glmmTMB(
  density ~ fyear + depth_c + depth_c2 + (1|region) + (1|X:Y),  # Station nested in region
  data = dat,
  family = tweedie(link = "log")
)

cat("✓ GLMM models fitted\n")
```

---

# STEP 4: Model Selection & Validation

## 4.1 AIC Comparison (All 12 Models)

```{r aic_comparison}
# Compile AIC table
aic_results <- data.frame(
  Model = c(paste0("GLM", 1:4), paste0("GAM", 1:4), paste0("GLMM", 1:4)),
  Family = rep(c("GLM", "GAM", "GLMM"), each = 4),
  AIC = c(
    AIC(glm1), AIC(glm2), AIC(glm3), AIC(glm4),
    AIC(gam1), AIC(gam2), AIC(gam3), AIC(gam4),
    AIC(glmm1), AIC(glmm2), AIC(glmm3), AIC(glmm4)
  )
) |>
  arrange(AIC) |>
  mutate(
    Delta_AIC = AIC - min(AIC),
    AIC_Weight = exp(-0.5 * Delta_AIC) / sum(exp(-0.5 * Delta_AIC)),
    Rank = row_number()
  )

kable(aic_results, digits = 2,
      caption = "AIC Comparison: All 12 Candidate Models")
```

```{r plot_aic, fig.height=6}
ggplot(aic_results, aes(x = reorder(Model, -AIC), y = AIC, fill = Family)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = min(aic_results$AIC) + 2, 
             linetype = "dashed", color = "red") +
  geom_hline(yintercept = min(aic_results$AIC) + 10,
             linetype = "dashed", color = "orange") +
  coord_flip() +
  labs(title = "Model Comparison by AIC",
       subtitle = "Red line = ΔAIC = 2, Orange = ΔAIC = 10",
       x = "Model", y = "AIC") +
  theme(legend.position = "top")
```

## 4.2 Select Best Models per Family

```{r select_best_per_family}
# Best GLM
best_glm_name <- aic_results |> filter(Family == "GLM") |> slice_min(AIC, n = 1) |> pull(Model)
best_glm <- get(tolower(best_glm_name))

# Best GAM
best_gam_name <- aic_results |> filter(Family == "GAM") |> slice_min(AIC, n = 1) |> pull(Model)
best_gam <- get(tolower(best_gam_name))

# Best GLMM
best_glmm_name <- aic_results |> filter(Family == "GLMM") |> slice_min(AIC, n = 1) |> pull(Model)
best_glmm <- get(tolower(best_glmm_name))

cat("\n=== Best Models per Family ===\n")
cat("Best GLM: ", best_glm_name, " (AIC =", round(AIC(best_glm), 2), ")\n")
cat("Best GAM: ", best_gam_name, " (AIC =", round(AIC(best_gam), 2), ")\n")
cat("Best GLMM:", best_glmm_name, "(AIC =", round(AIC(best_glmm), 2), ")\n")
```

## 4.3 Detailed Validation: Best GLMM

### 4.3.1 DHARMa Diagnostics

```{r dharma_glmm, fig.height=6}
sim_glmm <- simulateResiduals(best_glmm, n = 1000)
plot(sim_glmm, main = paste("DHARMa Diagnostics:", best_glmm_name))
```

### 4.3.2 Residuals by Covariates

```{r residuals_glmm, fig.height=8}
# Extract residuals
dat$resid_glmm <- residuals(best_glmm, type = "pearson")

p1 <- ggplot(dat, aes(x = fyear, y = resid_glmm)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "A) Residuals by Year", x = "Year", y = "Pearson Residuals") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(dat, aes(x = depth, y = resid_glmm)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "B) Residuals vs Depth", x = "Depth (m)", y = "Pearson Residuals")

p3 <- ggplot(dat, aes(x = region, y = resid_glmm)) +
  geom_boxplot(fill = "coral") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "C) Residuals by Region", x = "Region", y = "Pearson Residuals")

p4 <- ggplot(dat, aes(sample = resid_glmm)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "D) Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles")

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = paste("Residual Diagnostics:", best_glmm_name))
```

## 4.4 Detailed Validation: Best GAM

### 4.4.1 GAM Check

```{r gam_check, fig.height=8}
par(mfrow = c(2, 2))
gam.check(best_gam)
par(mfrow = c(1, 1))
```

### 4.4.2 Smooth Terms

```{r plot_smooths, fig.height=8}
par(mfrow = c(2, 2))
plot(best_gam, pages = 1, scheme = 2, shade = TRUE)
par(mfrow = c(1, 1))
```

### 4.4.3 Residuals

```{r residuals_gam, fig.height=8}
dat_gam <- best_gam$model
dat_gam$resid_gam <- residuals(best_gam, type = "pearson")

p1 <- ggplot(dat_gam, aes(x = fyear, y = resid_gam)) +
  geom_boxplot(fill = "lightgreen") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "A) Residuals by Year", x = "Year", y = "Pearson Residuals") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(dat_gam, aes(x = depth, y = resid_gam)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "B) Residuals vs Depth", x = "Depth (m)", y = "Pearson Residuals")

p3 <- ggplot(dat_gam, aes(x = X, y = Y, color = resid_gam)) +
  geom_point(size = 1, alpha = 0.5) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  coord_equal() +
  labs(title = "C) Spatial Residuals", x = "X (km)", y = "Y (km)")

p4 <- ggplot(dat_gam, aes(sample = resid_gam)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "D) Q-Q Plot")

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = paste("Residual Diagnostics:", best_gam_name))
```

---

# STEP 5: Prediction Grid Construction

## 5.1 Standardization Strategy

**Goal**: Create a prediction grid where covariates are held at **reference values** except year.

**Reference values**:

For **GLMM** (parametric depth):
- Depth → mean depth (centered = 0)
- Region → modal category

For **GAM** (spatial smooth):
- Depth → sequence covering range (for marginal effects)
- X, Y → grid covering survey area (for spatial integration)

## 5.2 Prediction Grid for GLMM

```{r pred_grid_glmm}

# Calculate reference values
ref_depth <- mean(dat$depth)
ref_depth_c <- (ref_depth - mean(dat$depth)) / sd(dat$depth)
ref_depth_c2 <- ref_depth_c^2
ref_region <- names(sort(table(dat$region), decreasing = TRUE))[1]

# Build prediction grid WITH ALL VARIABLES USED IN MODEL
pred_glmm <- expand.grid(
  year = sort(unique(dat$year)),
  depth_c = ref_depth_c,
  depth_c2 = ref_depth_c2,
  log_depth = log(ref_depth + 1), 
  region = ref_region,
  X = mean(dat$X),  
  Y = mean(dat$Y)
) |>
  mutate(
    fyear = factor(year, levels = levels(dat$fyear))
  )

# Now predict
pred_link_glmm <- predict(best_glmm, newdata = pred_glmm,
                          type = "link", se.fit = TRUE,re.form = NA)




kable(head(pred_glmm, 5), digits = 3,
      caption = "Prediction Grid for GLMM (First 5 rows)")
```

## 5.3 Prediction Grid for GAM (Spatial Integration)

```{r pred_grid_gam}
# Create spatial grid (50×50 resolution)
grid_res <- 50
X_seq <- seq(min(dat$X), max(dat$X), length.out = grid_res)
Y_seq <- seq(min(dat$Y), max(dat$Y), length.out = grid_res)

# Expand for all years
pred_gam <- expand.grid(
  X = X_seq,
  Y = Y_seq,
  depth = mean(dat$depth),  # Fix depth at mean
  year = sort(unique(dat$year))
) |>
  mutate(
    fyear = factor(year, levels = levels(dat$fyear)),
    region = ref_region  # Fixed region
  )

cat("GAM prediction grid dimensions:", nrow(pred_gam), "rows\n")
cat("Grid structure:", grid_res, "×", grid_res, "spatial points ×", 
    length(unique(dat$year)), "years\n")

kable(head(pred_gam, 5), digits = 2,
      caption = "Prediction Grid for GAM (First 5 rows)")
```

---

# STEP 6: Index Calculation

## 6.1 Predict with Best GLMM

```{r predict_glmm}

# Transform to response scale
pred_glmm <- pred_glmm |>
  mutate(
    fit_link = pred_link_glmm$fit,
    se_link = pred_link_glmm$se.fit,
    fit = exp(fit_link),
    lwr = exp(fit_link - 1.96 * se_link),
    upr = exp(fit_link + 1.96 * se_link)
  )

# Normalize to base year (2003)
base_glmm <- pred_glmm$fit[pred_glmm$year == min(pred_glmm$year)]

index_glmm <- pred_glmm |>
  mutate(
    index = fit / base_glmm,
    index_lwr = lwr / base_glmm,
    index_upr = upr / base_glmm,
    cv = (upr - lwr) / (2 * 1.96 * fit),
    model = best_glmm_name
  ) |>
  select(year, index, index_lwr, index_upr, cv, model)

kable(index_glmm, digits = 4,
      caption = paste("Standardized Index from", best_glmm_name))
```

## 6.2 Predict with Best GAM (Spatial Integration)

```{r predict_gam}
# Predict on link scale
pred_link_gam <- predict(best_gam, newdata = pred_gam,
                        type = "link", se.fit = TRUE)

# Transform to response scale
pred_gam <- pred_gam |>
  mutate(
    fit_link = pred_link_gam$fit,
    se_link = pred_link_gam$se.fit,
    fit = exp(fit_link),
    lwr = exp(fit_link - 1.96 * se_link),
    upr = exp(fit_link + 1.96 * se_link)
  )

# Integrate over spatial grid (sum densities, divide by grid points)
index_gam_raw <- pred_gam |>
  group_by(year) |>
  summarise(
    mean_fit = mean(fit),
    mean_lwr = mean(lwr),
    mean_upr = mean(upr),
    se_fit = sd(fit) / sqrt(n()),
    .groups = "drop"
  )

# Normalize to base year
base_gam <- index_gam_raw$mean_fit[index_gam_raw$year == min(index_gam_raw$year)]

index_gam <- index_gam_raw |>
  mutate(
    index = mean_fit / base_gam,
    index_lwr = mean_lwr / base_gam,
    index_upr = mean_upr / base_gam,
    cv = (mean_upr - mean_lwr) / (2 * 1.96 * mean_fit),
    model = best_gam_name
  ) |>
  select(year, index, index_lwr, index_upr, cv, model)

kable(index_gam, digits = 4,
      caption = paste("Standardized Index from", best_gam_name))
```

## 6.3 Normalize Nominal Index

```{r normalize_nominal}
base_nominal <- nominal_index$mean_density[nominal_index$year == min(nominal_index$year)]

index_nominal <- nominal_index |>
  mutate(
    index = mean_density / base_nominal,
    index_lwr = (mean_density - 1.96*se) / base_nominal,
    index_upr = (mean_density + 1.96*se) / base_nominal,
    cv = se / mean_density,
    model = "Nominal"
  ) |>
  select(year, index, index_lwr, index_upr, cv, model)
```

## 6.4 Compare All Three Indices


```{r compare_indices, fig.height=7}  
# Combine all indices
all_indices <- bind_rows(
  index_nominal,
  index_glmm,
  index_gam
)

# Reorder for better legend
all_indices <- all_indices |>
  mutate(model = factor(model, levels = c("Nominal", 
                                          unique(model)[unique(model) != "Nominal"])))

# Plot comparison
p=ggplot(all_indices, aes(x = year, y = index, color = model, fill = model)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_ribbon(aes(ymin = index_lwr, ymax = index_upr),
              alpha = 0.2, color = NA) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Abundance Index Comparison",
    subtitle = "Nominal vs Best GLMM vs Best GAM",
    x = "Year",
    y = "Relative Index (normalized to 2003)",
    color = "Model",
    fill = "Model"
  ) +
  scale_x_continuous(breaks = seq(2003, 2017, 2)) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

print(p)
```



```{r index_summary_table}
# Summary statistics
index_summary <- all_indices |>
  group_by(model) |>
  summarise(
    Mean_CV = mean(cv),
    Min_Index = min(index),
    Max_Index = max(index),
    Trend = (index[year == max(year)] - index[year == min(year)]) / index[year == min(year)],
    .groups = "drop"
  )

kable(index_summary, digits = 3,
      caption = "Index Summary Statistics by Model")
```

**Key Observations**:

- **Nominal index** shows high inter-annual variability, particularly in early years (2003-2006), reflecting changes in sampling locations and environmental conditions. Overall decline of ~X% from 2003 to 2017.

- **GLMM standardization** (with random effects for region and spatial stations `(1|region) + (1|X:Y)`) corrects for depth and spatial sampling heterogeneity, resulting in a smoother temporal trend. The standardized index shows a Y% decline, Z percentage points different from nominal, indicating that confounding factors (depth distribution, regional shifts) were masking/amplifying the true abundance trend.

- **GAM with spatial smooth** `s(X,Y, k=20)` additionally accounts for fine-scale spatial structure not captured by regional random effects. The GAM index shows [similar/slightly different] trend (W% correlation with GLMM), suggesting that the 2D spatial smooth captures residual spatial patterns beyond the region+station random effects.

- Uncertainty (CVs) are [compare]

---

# STEP 7: Final Diagnostics & Spatial Analysis

## 7.1 Spatial Autocorrelation in GLMM Residuals

```{r spatial_map_glmm, fig.height=7}
spatial_resid_glmm <- dat |>
  group_by(X, Y,fyear) |>
  summarise(mean_resid = mean(resid_glmm, na.rm = TRUE), .groups = "drop")

ggplot(spatial_resid_glmm, aes(x = X, y = Y, color = mean_resid)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-2, 2),
                       name = "Mean\nResidual") +
  coord_equal() +
  labs(title = paste("Spatial Residuals:", best_glmm_name),
       subtitle = "Clusters indicate unmodeled spatial structure",
       x = "X (km)", y = "Y (km)")


ggplot(spatial_resid_glmm, aes(x = X, y = Y, color = mean_resid)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-2, 2),
                       name = "Mean\nResidual") +
    facet_wrap(~fyear, ncol = 3) +
  coord_equal() +
  labs(title = paste("Spatial Residuals:", best_glmm_name),
       subtitle = "Clusters indicate unmodeled spatial structure",
       x = "X (km)", y = "Y (km)")


ggplot(dat, aes(x = X, y = Y, z = resid_glmm)) +
  stat_summary_2d(fun = mean, bins = 30, alpha = 0.8) +
  geom_contour(aes(color = after_stat(level)), bins = 10, linewidth = 0.5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-2, 2),
                       name = "Mean\nResidual") +
  facet_wrap(~fyear, ncol = 3) +
  coord_equal() +
  labs(title = paste("Spatial Residual Hotspots:", best_glmm_name),
       subtitle = "Smoothed interpolation reveals persistent spatial patterns",
       x = "X (km)", y = "Y (km)") +
  theme_minimal()

```



## 7.2 Spatial Autocorrelation in GAM Residuals


```{r spatial_map_gam, fig.height=7}
spatial_resid_gam <- dat_gam |>
  group_by(X, Y, fyear) |>
  summarise(mean_resid = mean(resid_gam, na.rm = TRUE), .groups = "drop")

ggplot(spatial_resid_gam, aes(x = X, y = Y, color = mean_resid)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-2, 2),
                       name = "Mean\nResidual") +
  coord_equal() +
  labs(title = paste("Spatial Residuals:", best_gam_name),
       subtitle = "GAM with s(X,Y) should reduce spatial patterns",
       x = "X (km)", y = "Y (km)")


ggplot(spatial_resid_gam, aes(x = X, y = Y, color = mean_resid)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-2, 2),
                       name = "Mean\nResidual") +facet_wrap(~fyear)+
  coord_equal() +
  labs(title = paste("Spatial Residuals:", best_gam_name),
       subtitle = "GAM with s(X,Y) should reduce spatial patterns",
       x = "X (km)", y = "Y (km)")


ggplot(dat_gam, aes(x = X, y = Y, z = resid_gam)) +  # z = resid_gam !
  stat_summary_2d(fun = mean, bins = 30, alpha = 0.8) +
  geom_contour(aes(z = resid_gam), color = "white", alpha = 0.5, bins = 10, linewidth = 0.5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-2, 2),
                       name = "Mean\nResidual") +
  facet_wrap(~fyear, ncol = 3) +
  coord_equal() +
  labs(title = paste("Spatial Residual Hotspots:", best_gam_name),
       subtitle = "GAM with s(X,Y) should eliminate spatial patterns",
       x = "X (km)", y = "Y (km)") +
  theme_minimal()


```


## 7.3 Transition to sdmTMB (If Needed)

```{r fit_sdmtmb}

  # Create mesh
  mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 5)

  # Fit spatiotemporal model
  fit_sdmtmb <- sdmTMB(
    density ~ 0+fyear + s(depth),
    data = dat,
    mesh = mesh,
    family = tweedie(link = "log"),
    spatial = "on",           # Spatial random field
    spatiotemporal = "iid",   # Independent spatiotemporal field
    time = "year"
  )

  # Sanity checks
  sanity(fit_sdmtmb)

  # Extract residuals
  dat$resid_sdmtmb <- residuals(fit_sdmtmb)

```

## 7.4 Compare Spatial Patterns Across Models

```{r compare_spatial_residuals, fig.height=12}
if(!is.null(fit_sdmtmb)) {

  # Prepare data for faceting
resid_comparison <- bind_rows(
  dat |> select(X, Y, fyear) |> mutate(resid = scale(dat$resid_glmm), model = best_glmm_name),
  dat_gam |> select(X, Y, fyear) |> mutate(resid = scale(dat_gam$resid_gam), model = best_gam_name),
  dat |> select(X, Y, fyear) |> mutate(resid = scale(dat$resid_sdmtmb), model = "sdmTMB")
)

  ggplot(resid_comparison, aes(x = X, y = Y, color = resid)) +
    geom_point(size = 1.5, alpha = 0.6) +
    facet_grid(model ~ fyear) +
    scale_color_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0, limits = c(-1.5, 1.5),
                         name = "Residual",oob = scales::squish) +
    coord_equal() +
    labs(title = "Spatial Residual Comparison Across Models",
         subtitle = "3 years × 3 models",
         x = "X (km)", y = "Y (km)") +
    theme(strip.text = element_text(size = 10, face = "bold"))
}
```

---

# Conclusions & Recommendations

 **For stock assessment**:
   - Provide index with 95% confidence intervals
   - Report CV by year
   - Include sensitivity analysis (compare models)
   - Document all modeling decisions
